Combining checkpointing and data compression\\ for high frequency seismic imaging

### 2018-10-03 paper/paper.tex
revised section 2

Compared to using only checkpointing without compression, our combined method
often improves performance. This is a consequence of the reduced size of stored
program states, allowing more program states to be stored during the forward
computation. This in turn reduces the amount of recomputation that needs to be
performed. On the other hand, the compression and decompression itself takes
time. We thereforen provide a comprehensive performance model that predicts
whether the combined method is beneficial, that is, whether the time spent
compressing and decompressing is less than the time saved by reduced
recomputations. We present various scenarios that use a TODO describe Richter
and TODO describe JLSE/SKX node system, error tolerances for the lossy ZFP
compression algorithm ranging from TODO to TODO, and different types of forward
and adjoint computations that vary in their ratio between compute cost and state
size.

@@

\section{Test case using Devito and pyRevolve}
We use Devito \cite{devito-compiler} \cite{devito-api} to setup wave equation solvers for forward and
adjoint modes. Devito is a domain-specific language to enable the rapid development of finite-difference
solvers from a high-level description of the partial differential equation. The simplest version of the wave
equation is the acoustic:
\begin{equation}
m(x)\frac{\partial^2 u(t, x)}{\partial t^2} - \Laplace u(t, x) = q(t, x)
\label{eqn:wave}
\end{equation}
where $m(x) = \frac{1}{c^2(x)}$ is the squared slowness with $c(x)$ the spatially dependent speed of sound, 
$u(t, x)$ is the wavefield, $\Laplace u(t, x)$ denotes the laplacian of the wavefield and $q(t, x)$ is the source.
Some of the kernels used in the performance measurements corresponded to a more physically representative
but complex version of this equation called Tilted Transverse Isotropy (TTI) \cite{zhang2011stable}. We leave the 
TTI equations out of this paper for brevity.

The solution to equation \ref{eqn:wave} form the forward problem. The complete optimization problem involves
the minimization of the misfit between the simulated signal and the observed signal as follows:

\begin{equation}
\min_{m} \phi_s(m) = \frac{1}{2} \left\lVert d_{sim} - d_{obs} \right\rVert_2^2
\end{equation}

This optimization problem is usually solved using gradient-descent methods and the gradient calculation is done
using the adjoint-state method, which involves the data-flow pattern from figure \ref{fig:dataflow}.

The values of $m(x)$ were derived from the SEAM Overthrust model \cite{aminzadeh1996three} which was over a grid
of $287 \times 881 \times 881$ points, including an absorbing layer of 40 points on each side. The grid spacing
was 25m in space and TODO in time. This was run for 2500 timesteps, and the final field stored as
representative of a typical snapshot. Figure \ref{fig:uncompressed} shows this stored field that was used for the compression tests.

@@

To implement Revolve with Devito, we used pyRevolve \cite{kukreja2018high} which is a python wrapper over the
original Revolve utility. The performance model that will be developed in section \ref{sec:performance_model} of
this paper assumes that the implementation is similar to pyRevolve, which stores a checkpoint by copying a portion
of the operator's working memory to the checkpointing memory and similarly loads a checkpoint by copying from the
checkpointing memory to the operator's working memory. Although a perfect implementation of checkpointing may
be able to avoid these copies, the overhead attached to these copies can be ignored for an operator that is
sufficiently computationally intensive. However, we include the overheads in the model to verify this assumption.

The experiments were run on two different systems, TODO description of Richter, henceforth called Richter and a
TODO description of skylake, henceforth called Skylake.

>>>>

Compared to using only checkpointing without compression, our combined method
often improves performance. This is a consequence of the reduced size of stored
program states, allowing more program states to be stored during the forward
computation. This in turn reduces the amount of recomputation that needs to be
performed. On the other hand, the compression and decompression itself takes
time. We thereforen provide a comprehensive performance model that predicts
whether the combined method is beneficial, that is, whether the time spent
compressing and decompressing is less than the time saved by reduced
recomputations.

We present test cases that use a TODO describe Richter
and TODO describe JLSE/SKX node system, henceforth called Richter and Skylake,
respectively. We discuss performance results for varying error tolerances for
the lossy ZFP compression algorithm ranging from TODO to TODO, and different
types of forward and adjoint computations that vary in their ratio between
compute cost and state size.

@@

\section{Test case using Devito and pyRevolve}
We use Devito~\cite{devito-api,devito-compiler} to solve forward and adjoint
wave equation problems. Devito is a domain-specific language to enable the
rapid development of finite-difference solvers from a high-level description
of a partial differential equation. The simplest version of the wave
equation is the acoustic wave equation given by
\begin{equation}
m(x)\frac{\partial^2 u(t, x)}{\partial t^2} - \Laplace u(t, x) = q(t, x),
\label{eqn:wave}
\end{equation}
where $m(x) = \frac{1}{c^2(x)}$ is the squared slowness, $c(x)$ the spatially
dependent speed of sound, $u(t, x)$ is the wavefield, $\Laplace u(t, x)$
denotes the laplacian of the wavefield and $q(t, x)$ is a source term.
Some of our experiments in Section TODO are performed using a more accurate
and more complex version of this equation called Tilted Transverse Isotropy
(TTI)~\cite{zhang2011stable}. We leave the TTI equations out of this paper
for brevity.

The solution to equation~\ref{eqn:wave} forms the forward problem. The adjoint
problem minimizes the misfit between simulated and observed signal given by
\begin{equation}
\min_{m} \phi_s(m) = \frac{1}{2} \left\lVert d_{sim} - d_{obs} \right\rVert_2^2.
\end{equation}
This optimization problem is usually solved using gradient-descent methods,
where the gradient is computed using the adjoint-state method, which involves
the data-flow pattern from figure \ref{fig:dataflow}.

The values of $m(x)$ are in this work derived from the SEAM Overthrust
model~\cite{aminzadeh1996three} over a grid of $287 \times 881 \times 881$
points, including an absorbing layer of 40 points on each side. The grid spacing
is 25m in space and TODO in time. This is run for 2500 timesteps. The wave field
at the final time is shown in Figure~\ref{fig:uncompressed}, and only this field
was used for the compression experiments in this paper.

@@

To implement Revolve with Devito, we use pyRevolve~\cite{kukreja2018high} which
is a python wrapper for the Revolve algorithm. The performance model in
section~\ref{sec:performance_model} assumes that the implementation is similar
to pyRevolve, which stores a checkpoint by copying a portion of the operator's
working memory to the checkpointing memory and similarly loads a checkpoint by
copying from the checkpointing memory to the operator's working memory. Although
a perfect implementation of checkpointing may be able to avoid these copies, the
overhead attached to these copies can be ignored for an operator that is
sufficiently computationally intensive. However, we include the overheads in the
model to verify this assumption.
